---
title: 'A One-Stop Shop for Heritability Estimation'
author: Fonti Kar and Emi Tanaka
date: '2026-01-12'
editor_options: 
  chunk_output_type: console
---

## What is heritability?
In genetics, heritability quantifies how much variation in a trait is due to genetic differences vs. environmental effects. It provides an indication of how evolvable traits are which is important for fields such as evolutionary ecology, quantitative genetics and breeding research. 

## The challenge
Recently, a number of statistical methods have been proposed for generalising heritability to account for factors such as unbalanced replication. These competing methods are important for accounting different sources of variation, but they have not been implemented in an accessible way. 

To complicate matters further, its not common convention to openly report on _how_ we calculate heritability. Rarely do we share our heritability code scripts either. This makes it difficult to reproduce each others calculations and objectively make interpretations across studies. 

### 📦 Introducing {heritable}

`heritable` is an R package designed to compute a variety of heritability estimates from mixed model outputs. 
Our goal is to provide a centralised toolkit for heritability calculations. Currently, we have implemented six  **broad-sense heritability** methods for **single environment models**.  Our functions are compatible with mixed model objects from packages `lme4` and `asreml`

You can download `heritable` from CRAN! 

```{r}
# install.packages("heritable")

library(heritable)
```

Alternatively, if you would like the latest developements you can download `heritable` directly from our Github repo

```{r}
# remotes::install_github("anu-aagi/heritable")
```

#### Heritability methods 

We have implemented the following for broad-sense heritability: 

- [Cullis' method](https://link.springer.com/content/pdf/10.1198%2F108571106X154443.pdf){target="_blank"}
- [Oakey's method](link.springer.com/content/pdf/10.1007%2Fs00122-006-0333-z.pdf){target="_blank"}
- [Piepho's method](https://academic.oup.com/genetics/article-pdf/177/3/1881/37411981/genetics1881.pdf){target="_blank"}
- [Delta method](https://academic.oup.com/genetics/article/212/4/991/5931331){target="_blank"}
- Standard method

You can learn more about these on [our package website](https://anu-aagi.github.io/heritable/reference/index.html){target="_blank"}  

These methods come from mixed model theory and differ in how to obtain the necessary variance components for calculations. We recommend having a good read of [Paul Schmidt's and colleagues review paper](https://academic.oup.com/genetics/article/212/4/991/5931331){target="_blank"} to learn more about their derivation and nuances. 

## A quick demo

We'll use the `lettuce` dataset included in the package to show you the workflow of the package

### Fit a single environment model

Let's fit a single environment model using `lme4` and `asreml`, noting that the latter requires a license to run. 

```{r}
#| message: false
#| warning: false
library(lme4)
fit_lme4 <- lmer(y ~ rep + (1|gen) + (1|rep), data = lettuce_phenotypes)
```

```{r}
#| message: false
#| warning: false
library(asreml)
fit_asreml <- asreml(
  fixed = y ~ rep,
  random = ~ gen,
  data = lettuce_phenotypes,
  trace = FALSE
)
```

### Calculate broad-sense heritability

By default, the wrapper `H2()` will compute all available estimates for you. 

```{r}
H2(fit_lme4, target = "gen")
H2(fit_asreml, target = "gen")
```

## What's next for {heritable} and contributing to the package

We have a few features in the pipeline including: 

- Narrow-sense heritability calculations
- Confidence intervals for estimates
- Estimates for G x E models
- Expand compatibility with `glmmTMB` and Bayesian models `brms` or `rstan`
- Data visualiation of variance decomposition

We are always excited to receive feedback, hear new ideas and find ways to collaborate. Get in touch by submitting a [Github issue at our repository](https://github.com/anu-aagi/heritable){target="_blank"}! 

## Learn more

Check out Fonti's [Biometrics in the Bush Capital conference talk](https://biometricsociety.org.au/conference2025/slides/Contributed/S11B1-Fonti%20Kar.pdf), where she soft launches `heritable`